\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-12-09 manuscript.tex
Edits
reformatting
edits throughout for PeerJ

\author[1]{Laura Vann}%author information
\author[1,2]{Thomas Kono}
\author[1,3]{Tanja Pyh\"aj\"arvi}
\author[1,4,6]{Matthew B. Hufford}
\author[1,5,6]{Jeffrey Ross-Ibarra}
\affil[1]{Department of Plant Sciences, University of California Davis, Davis, CA, USA}
\affil[2]{Department of Agronomy and Plant Genetics, University of Minnesota Twin Cities, Minneapolis, MN, USA}
\affil[3]{Department of Biology, University of Oulu, Oulu, Finland}
\affil[4]{Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, USA}
\affil[5]{Center for Population Biology and Genome Center, University of California Davis, Davis, CA, USA}
\affil[6]{Author for Correspondence}
\date{}

@@

\begin{linenumbers}
	\pagewiselinenumbers
	\renewcommand\linenumberfont{\normalfont\bfseries\small}

\clearpage%inserts page break
%\footnotetext{\flushleft{Manuscript received \line(1,0){120};   revision accepted \line(1,0){120}.}
\vspace*{1cm}
%\subsection*{Acknowledgements}%subsection header
The authors thank the Department of Plant Sciences at UC Davis for graduate student research funding to LEV and for research funds supporting the project, UC Mexus for a postdoctoral scholar grant to MBH and JR-I, and G. Coop for helpful discussion.
}
\clearpage

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as the `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. Often, after identification, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovachMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, we know almost nothing about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1}, and specifically the domesticated haplotype at \emph{tb1}, in natural populations of the wild ancestor of domesticated maize.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes of \emph{tb1}\citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of \emph{parviglumis}, \emph{mexicana}, and landrace maize; however little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations.

@@

\begin{centering}
\section*{MATERIALS AND METHODS}
\end{centering}

\textbf{Sampling and genotyping---}We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Numbers of individuals sampled ranged from 1-43 for \emph{parviglumis}, 1-35 for \emph{mexicana}, and 1-18 for the maize landrace populations. DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion due to its large size ($\sim$5kb). Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \url{http://dx.doi.org/10.6084/m9.figshare.1165577}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, with the assumption being that individuals fall into either a maize or teosinte cluster, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency within populations in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled sixteen or more chromosomes. We used 16 as a cut-off for the minimum number of chromosomes sampled from a population. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to genome-wide estimates from previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is closer to zero in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

\textbf{Evidence of introgression---}The highest frequency of the \emph{Hopscotch} insertion in teosinte was found in \emph{parviglumis} sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the \emph{Hopscotch} element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome one and specifically in the \emph{tb1} region. If the \emph{Hopscotch} is found in these populations due to recent introgression we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the \emph{Hopscotch} and SNPs surrounding the \emph{tb1} region in our resequenced populations (Figure~\ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region  does not differ significantly between populations with (average $r^{2}$ of 0.085) and without (average $r^{2}=0.082$) the \emph{Hopscotch} insertion. In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$; Table \ref{Table3R2}).

@@

\textbf{Phenotyping---}To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion based on our initial genotyping. 
We measured tiller number and tillering index, the ratio of the sum of tiller lengths to plant height, for 206 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. 
We also measured culm diameter, a phenotype that differs between maize and teosinte but has not been shown to be affected by the \emph{Hopscotch} insertion \citep{Briggs2007}. Culm diameter is meant to be an independent trait against which we can compare patterns of tillering index x \emph{Hopscotch} genotype data. If tillering index in \emph{parviglumis} is affected by the \emph{Hopscotch} insertion the expectation is that patterns of tillering index data will have a significant correlation with \emph{Hopscotch} genotype, whereas we should find no significant correlation between culm diameter and \emph{Hopscotch} genotype. 
Phenotypic data are available at \url{http://dx.doi.org/10.6084/m9.figshare.776926}. 
Our plantings produced 82 homozygotes for the \emph{Hopscotch} insertion at \emph{tb1}, 104 heterozygotes, and 20 homozygotes lacking the insertion; these numbers do not deviate from expectations of Hardy-Weinberg equilibrium. 
After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find no significant correlation between genotype at the \emph{Hopscotch} insertion and tillering index (Fig. \ref{Fig4Boxplots}), tiller number, or culm diameter. 
Only on day 40 did we observe a weak but statistically insignificant ($r^2= 0.02$, $p=0.0848$) correlation between tillering index and the \emph{Hopscotch} genotype, although in the opposite direction of that expected, with homozygotes for the insertion showing a higher tillering index.

@@

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings. Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster. Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations. 
Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in populations of \emph{parviglumis}. Overall nucleotide diversity is 76\% less than seen in the sequences from the 66kb upstream region, and Tajima's D is considerably lower and consistently negative across populations (Table \ref{Table2Diversity}). In fact, values of Tajima's D in the 5' UTR are toward the extreme negative end of the distribution of this statistic previously calculated across loci sequenced in \emph{parviglumis} \citep{Wright27052005, Moeller01072007}. Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in Jaliscan populations of \emph{parviglumis}.

@@

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is widespread in populations of \emph{parviglumis} and \emph{mexicana} and occasionally at high allele frequencies. 
Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression, and it seems unlikely that the insertion would have drifted to high frequency in multiple populations. 
We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}.  Coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations, this suggests that the locus may play an ecological role in teosinte.

@@

\bibliography{refs}%print bibliography

@@

%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
\begin{center}
   \includegraphics[width=0.6\textwidth]{Fig1MapBalsas.pdf}
   \end{center}
    \caption{Map showing the frequency of the \emph{Hopscotch} allele in populations of \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of alleles sampled. The Balsas River is shown, as the Balsas River Basin is believed to be the center of domestication of maize.} 
\label{Fig1Map}
\end{figure*}
%--------------------------------------------
\clearpage

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=140mm]{Fig2LDPlots.pdf}
   \end{center}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D) San Lorenzo. The upper triangle above the black diagonal is colored based on the $r^{2}$ value between SNPs while the bottom triangle is colored based on p-value for the corresponding  $r^{2}$ value.}
\label{Fig2LD}
\end{figure*}
%-------------------------------------------------------------------
\clearpage

>>>>

\author[1]{Authors: Laura Vann}%author information
\author[1,2]{Thomas Kono}
\author[1,3]{Tanja Pyh\"aj\"arvi}
\author[1,4,6]{Matthew B. Hufford}
\author[1,5,7]{Jeffrey Ross-Ibarra}
\affil[1]{Department of Plant Sciences, University of California Davis, Davis, CA, USA}
\affil[2]{Department of Agronomy and Plant Genetics, University of Minnesota Twin Cities, Minneapolis, MN, USA}
\affil[3]{Department of Biology, University of Oulu, Oulu, Finland}
\affil[4]{Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, USA}
\affil[5]{Center for Population Biology and Genome Center, University of California Davis, Davis, CA, USA}
\affil[6]{Corresponding Author: Matthew B. Hufford; 339A Bessey Hall, Iowa State University, Ames, IA, USA; phone: 1-515-294-8511; email: mhufford@iastate.edu}
\affil[7]{Corresponding Author: Jeffrey Ross-Ibarra; 262 Robbins Hall, Mail Stop 4, University of California, Davis, CA, USA; phone: 1-530-752-1152; email: rossibarra@ucdavis.edu}
\date{}

@@

\begin{linenumbers}
	\renewcommand\linenumberfont{\normalfont\bfseries\small}

\clearpage%inserts page break
%\footnotetext{\flushleft{Manuscript received \line(1,0){120};   revision accepted \line(1,0){120}.}
\vspace*{1cm}

@@

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. 
Plant domestication usually involves a suite of phenotypic changes such as loss of seed shattering and increased fruit or grain size, which are commonly referred to as the `domestication syndrome' \citep{OlsenWendel2013}, and much of the study of domestication has focused on understanding the genetic variation underlying these traits \citep{OlsenGross2010}. 
Because most domesticates show reduced genetic diversity relative to their wild counterparts, effort has been made to identify agronomically useful variation in crop wild relatives \citep{flint2009wide}. In some instances, the alleles conferring these beneficial traits are bred into domesticates for crop improvement. For example, \emph{Oryza rufipogon}, the wild progenitor of domesticated rice, has proven useful for the integration of a number of beneficial QTL controlling traits such as grain size and yield into domesticated rice \citep{KovachMcCouch2008}. In addition to researching the role of wild alleles in domesticates, researchers have also investigated the role of variation in domesticated taxa in the evolution of feral and weedy populations \citep{ellstrand2010crops}. But even though domesticated alleles are often found segregating in wild relatives \citep{Gallavotti2004, SigmonVollbrecht2010}, little is known about the ecological role of this variation in natural populations. In this paper we present an ecological genetic analysis of the domestication locus \emph{tb1}---specifically the domesticated haplotype at \emph{tb1}---in natural populations of the wild ancestor of domesticated maize.

@@

Many morphological changes are associated with maize domestication, and understanding the genetic basis of these changes has been a focus of maize research for a number of years \citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in  branching \citep{DoebleyStecGustus1995} and tillering \citep{DoebleyStec1991} differences between maize and teosinte.  A 4.9 kb retrotransposon (\emph{Hopscotch}) insertion into the upstream control region of \emph{tb1} in maize acts to enhance expression of \emph{tb1}, thus repressing lateral organ growth \citep{Doebley1997, Studer2011}.  Dating of the \emph{Hopscotch} retrotransposon suggests that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. The effects of the \emph{Hopscotch} insertion have been studied in maize \citep{Studer2011}, and analysis of teosinte alleles at \emph{tb1} has identified functionally distinct allelic classes of \emph{tb1}\citep{StuderDoebley2012}, but little is known about the role of \emph{tb1} or the \emph{Hopscotch} insertion at this locus in natural populations of teosinte. Previous studies have confirmed the presence of the \emph{Hopscotch} in samples of \emph{parviglumis} and landrace maize \citep{Studer2011}; however, little is known about the frequency with which the \emph{Hopscotch} is segregating in natural populations.

@@

\begin{centering}
\section*{MATERIALS \& METHODS}
\end{centering}

\textbf{Sampling and genotyping---}We sampled 1,110 individuals from 350 accessions (247 maize landraces, 17 \emph{mexicana} populations, and 86 \emph{parviglumis} populations) and assessed the presence or absence of the \emph{Hopscotch} insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Numbers of individuals sampled ranged from 1-43 for \emph{parviglumis}, 1-35 for \emph{mexicana}, and 1-18 for the maize landrace populations. DNA was extracted from leaf tissue using a modified CTAB approach \citep{DoyleDoyle1990, Maloof1984}. We designed primers using PRIMER3 \citep{RozenSkaletsky2000} implemented in Geneious \citep{Kearse2012} to amplify the entire \emph{Hopscotch} element, as well as an internal primer allowing us to simultaneously check for possible PCR bias between presence and absence of the \emph{Hopscotch} insertion due to its large size ($\sim$5kb). Two PCRs were performed for each individual, one with primers flanking the \emph{Hopscotch} (HopF/HopR) and one with a flanking primer and an internal primer (HopF/HopIntR). Primer sequences are HopF, {\small 5'-TCGTTGATGCTTTGATGGATGG-3'}; 
HopR, {\small 5'-AACAGTATGATTTCATGGGACCG-3'}; and HopIntR, {\small  5'-CCTCCACCTCTCATGAGATCC-3'} (Appendix 3 and Appendix 4, See Supplemental Materials with the online version of this article). Homozygotes show a single band for absence of the element ($\sim$300bp) and two bands for presence of the element ($\sim$5kb, amplification of the entire element, and $\sim$1.1kb, amplification of part of the element), whereas heterozygotes show all three bands (Appendix 2, See Supplemental Materials with the online version of this article). Since we developed a PCR protocol for each allele, if only one PCR resolved well, we scored one allele for that individual rather than infer the diploid genotype. We used Phusion High Fidelity Enzyme (Thermo Fisher Scientific Inc., Waltham, Massachusetts, USA) and the following conditions for amplifications: 98$^{\circ}$C for 3 min, 30 cycles of 98$^{\circ}$C for 15 s, 65$^{\circ}$C for 30 s, and 72$^{\circ}$C for 3 min 30 s, with a final extension of 72$^{\circ}$C for 10 min. PCR products were visualized on a 1\% agarose gel and scored for presence/absence of the \emph{Hopscotch} based on band size.

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \citep{Falush2003} and phased data from \citet{Pyhajarvi2013}, combined with the corresponding SNP data from a diverse panel of 282 maize lines \citep{Cook2012}. SNPs were anchored in a modified version of the IBM genetic map \citep{Gerke2013}. Since STRUCTURE does not account for LD due to physical linkage we created haplotype blocks using a custom Perl script from \citet[][code available at \url{http://dx.doi.org/10.6084/m9.figshare.1165577}]{Hufford2013}.  In maize, LD decays over an average distance of 5500bp \citep{Chia2012}; because LD decay is even more rapid in teosinte \citep{Chia2012} we used a conservative haplotype block size of 5kb.  We ran STRUCTURE at K=2 under the linkage model, with the assumption being that individuals fall into either a maize or teosinte cluster, performing three replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

\textbf{Genotyping---}The genotype at the \emph{Hopscotch} insertion was confirmed with two PCRs for 837 individuals of the 1,100 screened. Among the 247 maize landrace accessions genotyped, all but eight were homozygous for the presence of the insertion (Appendix 1 and Appendix 2, See Supplemental Materials with the online version of this article). Within our \emph{parviglumis} and \emph{mexicana} samples we found the \emph{Hopscotch} insertion segregating in 37 ($n=86$) and four ($n=17$) populations, respectively, and at highest frequency within populations in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico (Fig. \ref{Fig1Map}). Using our \emph{Hopscotch} genotyping, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled sixteen or more chromosomes. We found that F\textsubscript{CT} = 0, and levels of F\textsubscript{ST} among populations within each subspecies (0.22) and among all populations (0.23) are similar to genome-wide estimates from previous studies (\citealt{Pyhajarvi2013}; Table \ref{Table1Fst}). Although we found large variation in \emph{Hopscotch} allele frequency among our populations, BayEnv analysis did not indicate a correlation between the \emph{Hopscotch} insertion and environmental variables (all Bayes Factors $<$ 1).

\textbf{Sequencing---}To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the \emph{tb1} region and any evidence of selection on this locus, we sequenced two small ($<$1kb) regions upstream of the \emph{tb1} ORF in four populations. After alignment and singleton checking we recovered 48 and 40 segregating sites for the 5' UTR region (Region 1) and the 66kb upstream region (Region 2), respectively. For Region 1, Ejutla A has the highest values of haplotype diversity and $\theta_\pi$, while Ejutla B and La Mesa have comparable values of these summary statistics, and San Lorenzo has much lower values. Additionally, Tajima's D is strongly negative in the two Ejutla populations and La Mesa, but is closer to zero in San Lorenzo (Table \ref{Table2Diversity}, Appendix 2, See Supplemental Materials with the online version of this article). For Region 2, haplotype diversity and $\theta_\pi$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table \ref{Table2Diversity}). Tajima's D is positive in all populations except La Mesa, indicating an excess of low frequency variants in this population (Table \ref{Table2Diversity}). Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are close to zero for both sequenced regions as well as for the \emph{Hopscotch}, while they are high for other population pairs (Table \ref{Table1Fst}). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILs; data from Maize HapMapV2, \citealt{Chia2012}) do not reveal any clear clustering pattern with respect to population or \emph{Hopscotch} genotype (Appendix 5, See Supplemental Materials with the online version of this article); individuals within our sample that have the \emph{Hopscotch} insertion do not group with the teosinte inbred lines or domesticated maize that have the \emph{Hopscotch} insertion.

\textbf{Evidence of introgression---}The highest frequency of the \emph{Hopscotch} insertion in teosinte was found in \emph{parviglumis} sympatric with cultivated maize. Our initial hypothesis was that the high frequency of the \emph{Hopscotch} element in these populations could be attributed to introgression from maize into teosinte. To investigate this possibility we examined overall patterns of linkage disequilibrium across chromosome 1 and specifically in the \emph{tb1} region. If the \emph{Hopscotch} is found in these populations due to recent introgression from maize we would expect to find large blocks of linked markers near this element. We find no evidence of elevated linkage disequilibrium between the \emph{Hopscotch} and SNPs surrounding the \emph{tb1} region in our resequenced populations (Figure~\ref{Fig2LD}), and $r^{2}$ in the \emph{tb1} region  does not differ significantly between populations with (average $r^{2}$ of 0.085) and without (average $r^{2}=0.082$) the \emph{Hopscotch} insertion. In fact, average $r^{2}$ is lower in the \emph{tb1} region ($r^{2}=0.056$) than across the rest of chromosome 1 ($r^{2}=0.083$; Table \ref{Table3R2}).

@@

\textbf{Phenotyping---}To assess the contribution of \emph{tb1} to phenotypic variation in tillering in a natural population, we grew plants from seed sampled from the San Lorenzo population of \emph{parviglumis}, which had a high mean frequency (0.44) of the \emph{Hopscotch} insertion based on our initial genotyping. 
We measured tiller number and tillering index, the ratio of the sum of tiller lengths to plant height, for 206 plants from within the San Lorenzo population, and genotyped plants for the \emph{Hopscotch} insertion. 
We also measured culm diameter, a phenotype that differs between maize and teosinte but has not been shown to be affected by the \emph{Hopscotch} insertion \citep{Briggs2007}. Culm diameter is meant to be an independent trait against which we can compare patterns of tillering index x \emph{Hopscotch} genotype data. If tillering index in \emph{parviglumis} is affected by the \emph{Hopscotch} insertion, the expectation is that patterns of tillering index data will have a significant correlation with \emph{Hopscotch} genotype, whereas we should find no significant correlation between culm diameter and \emph{Hopscotch} genotype. 
Phenotypic data are available at \url{http://dx.doi.org/10.6084/m9.figshare.776926}. 
Our plantings produced 82 homozygotes for the \emph{Hopscotch} insertion at \emph{tb1}, 104 heterozygotes, and 20 homozygotes lacking the insertion; these numbers do not deviate from expectations of Hardy-Weinberg equilibrium. 
After performing a repeated measures ANOVA between our transformed tillering index data and \emph{Hopscotch} genotype we find no significant correlation between genotype at the \emph{Hopscotch} insertion and tillering index (Fig. \ref{Fig4Boxplots}), tiller number, or culm diameter. 
Only on day 40 did we observe a weak but statistically insignificant ($r^2= 0.02$, $p=0.0848$) correlation between tillering index and the \emph{Hopscotch} genotype, although in the opposite direction of that expected, with homozygotes for the insertion showing a higher tillering index.

@@

Remaining explanations for differential frequencies of the \emph{Hopscotch} among teosinte populations include both genetic drift and natural selection.  Previous studies using both SSRs and genome-wide SNP data have found evidence for a population bottleneck in the San Lorenzo population \citep{Hufford2010, Pyhajarvi2013}, and the lower levels of sequence diversity in this population in the 5' UTR (Region 1) coupled with more positive values of Tajima's D are consistent with these earlier findings. Such population bottlenecks can exaggerate the effects of genetic drift through which the \emph{Hopscotch} allele may have risen to high frequency entirely by chance. A bottleneck in San Lorenzo, however, does not explain the high frequency of the \emph{Hopscotch} in multiple populations in the Jalisco cluster. Moreover, available information on diversity and population structure among Jaliscan populations \citep{Hufford2010, Pyhajarvi2013} is not suggestive of recent colonization or other demographic events that would predict a high frequency of the allele across populations. 
Finally, diversity values in the 5' UTR of \emph{tb1} are suggestive of natural selection acting upon the gene in populations of \emph{parviglumis}. Overall nucleotide diversity is 76\% less than seen in the sequences from the 66kb upstream region, and Tajima's D is considerably lower and consistently negative across populations (Table \ref{Table2Diversity}). In fact, values of Tajima's D in the 5' UTR are toward the extreme negative end of the distribution of this statistic previously calculated across loci sequenced in \emph{parviglumis} \citep{Wright27052005, Moeller01072007}. Though not definitive, these results are consistent with the action of selection on the upstream region of \emph{tb1}, perhaps suggesting an ecological role for the gene in Jaliscan populations of \emph{parviglumis}.  Finally, while these results are consistent with selection at the \emph{tb1} locus in teosinte, they do not confirm selection specifically on the \emph{Hopscotch} insertion at this locus.

@@

\begin{centering}
\section*{CONCLUSIONS}
\end{centering}

In conclusion, our findings demonstrate that the \emph{Hopscotch} allele is widespread in populations of \emph{parviglumis} and \emph{mexicana} and occasionally at high allele frequencies. 
Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression, and it seems unlikely that the insertion would have drifted to high frequency in multiple populations. 
We do, however, find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}.  Coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations, this suggests that the locus---and potentially the domestication allele at this locus---may play an ecological role in teosinte.

@@

\subsection*{Acknowledgements}%subsection header
The authors thank Graham Coop for helpful discussion.

\clearpage
\bibliography{refs}%print bibliography

@@

%Figure 1 Map 
%--------------------------------------------
\begin{figure*}[!t]
\begin{center}
   \includegraphics[width=0.8\textwidth]{Fig1MapBalsas.png}
   \end{center}
    \caption{Map showing the frequency of the \emph{Hopscotch} allele in populations of \emph{parviglumis} where we sampled more than 6 individuals. Size of circles reflects number of individuals sampled. The Balsas River is shown, as the Balsas River Basin is believed to be the center of domestication of maize.} 
\label{Fig1Map}
\end{figure*}
%--------------------------------------------
\clearpage

% Figure 2. LD plots
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=180mm]{Fig2LDPlots.png}
   \end{center}
    \caption{Linkage disequilibrium for SNPs in Mb 261-268 on chromosome 1. The yellow rectangle indicates the location of the \emph{Hopscotch} insertion and the green rectangle represents the \emph{tb1} ORF. A) Ejutla A; B) Ejutla B; C) La Mesa; D) San Lorenzo. The upper triangle above the black diagonal is colored based on the $r^{2}$ value between SNPs while the bottom triangle is colored based on p-value for the corresponding  $r^{2}$ value.}
\label{Fig2LD}
\end{figure*}
%-------------------------------------------------------------------
\clearpage

% Figure 3 STRUCTURE Plot 4 Pops
%-------------------------------------------------------------------
\begin{figure*}[!t]
  \begin{center}
   \includegraphics[width=160mm]{Fig3Structure.png}
     \end{center}
    \caption{STRUCTURE assignment to maize across a section of chromosome 1. The dotted lines mark the beginning of the sequenced region 66kb upstream (Region 2) and the end of the \emph{tb1} ORF.}
\label{Fig3Structure}
\end{figure*}
%-------------------------------------------------------------------
\clearpage
